Approximation as Differentiation

### 2025-06-03 main/higher-order.tex
Notation for structure map and catamorphism for initial F-alg.
C bicartesian closed.

We introduce a standard total functional languages with a (reasonably) expressive type
system~\cite{crole94,pitts01,santocanale02}.

@@

Suppose $\cat{C}$ a Cartesian closed category with exponential $\internalHom{X}{Y}$, evaluation morphisms
$\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, and the following additional structure:
\begin{enumerate}
\item finite coproducts $(\coprod, \Zero)$;
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
\item $\mu$-polynomial endofunctors given by $\muPoly_{\cat{C}}$.
\end{enumerate}
\figref{default-semantics:types} gives the interpretation of a type $\tau$ with a single free type variable as
a $\mu$-polynomial endofunctor on $\cat{C}$.

>>>>

We introduce a standard total functional language with a (reasonably) expressive type
system~\cite{crole94,pitts01,santocanale02}, over a set $\PrimTy$ of primitive types and a set $\PrimOp$

@@

For any $\mu$-polynomial endofunctor $F: \cat{C} \to \cat{C}$ with initial algebra $\mu F$, write $\inMap_F$
for the structure map $F\mu F \to \mu F$ and write $\cata_f$ for the unique morphism $\mu F \to X$ from the
initial $F$-algebra to any $F$-algebra $(X, f: FX \to X)$.

For a ``default'' (baseline) semantics, fix a bicartesian closed category $\cat{C}$ with finite products
$(\times, \One)$, finite coproducts $(\coprod, \Zero)$, exponentials $\internalHom{X}{Y}$, evaluation
morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, and the following additional structure:
\begin{enumerate}
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
\item $\mu$-polynomial endofunctors given by $\muPoly_{\cat{C}}$.
\end{enumerate}
\figref{default-semantics:types} gives the default interpretation of types $\Delta \vdash \tau$ with
$\length{\Delta} = n$ as multifunctors $\cat{C}^n \to \cat{C}$ (or as objects of $\cat{C}$ in the case $n =
0$). \figref{default-semantics:terms} gives the default interpretation of terms $\Gamma \vdash t: \tau$ as
morphisms $\sem{\Gamma} \to \sem{\tau}$.
